---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
```{r}
library(ggplot2)
library(ggsci)
library(hdf5r)
library(patchwork)
library(RColorBrewer)
library(Seurat)
library(tidyverse)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, error = FALSE) 
```


```{r}
setwd("/Users/christinacomo/OneDrive - The University of Colorado Denver/Spatial")
getwd()
```
```{r}
sample <- "control1"
data_dir <- paste('/Users/christinacomo/OneDrive - The University of Colorado Denver/Spatial/data/control1/spatial', sample, sep = "")
so <- Load10X_Spatial(data_dir,
                      slice = sample)
```


```{r}
list.files(data_dir) # Should show filtered_feature_bc_matrix.h5
```


```{r}
SampleA <- Load10X_Spatial(
  data_dir,
  filename = "filtered_feature_bc_matrix.h5",
  assay = "Spatial",
  slice = "slice1",
  filter.matrix = TRUE,
  to.upper = FALSE,
  image = NULL,
)
```

```{r}
sample <- "control1"
SampleA[["orig.ident"]] <- sample
SampleA <- SetIdent(SampleA, value = so@meta.data$orig.ident)
```

```{r}
SampleA <- PercentageFeatureSet(SampleA, "^mt-", col.name = "percent_mito")
SampleA <- PercentageFeatureSet(SampleA, "^Hb.*-", col.name = "percent_hb")
```

```{r, fig.height = 3, fig.width = 3}
VlnPlot(SampleA, 
        features = c("nCount_Spatial", 
                     "nFeature_Spatial", 
                     "percent_mito", 
                     "percent_hb"),
        pt.size = 0.1, 
        ncol = 2) + NoLegend()
```

```{r}
SpatialFeaturePlot(SampleA, 
                   features = c("nCount_Spatial", 
                                "nFeature_Spatial", 
                                "percent_mito", 
                                "percent_hb"))
```

```{r}
SampleA <- SCTransform(SampleA, assay = "Spatial")
```

```{r}
SampleA <- RunPCA(SampleA, assay = "SCT", verbose = FALSE)
SampleA <- FindNeighbors(SampleA, reduction = "pca", dims = 1:30, verbose = FALSE)
SampleA <- FindClusters(SampleA, verbose = FALSE)
SampleA <- RunUMAP(SampleA, reduction = "pca", dims = 1:30)
```

```{r}
p1 <- DimPlot(SampleA, reduction = "umap", label = TRUE)
p2 <- SpatialDimPlot(SampleA, label = TRUE, label.size = 3)
p1 + p2
```

```{r}
saveRDS(SampleA, "/Users/christinacomo/Desktop/Spatial/RDS_files/SampleA_initial.RDS")
```


```{r}
#Now subset with  spots 
MGE <- read.csv("/Users/christinacomo/Desktop/Spatial/Raw Data/Control1MGE.csv")
LGE <- read.csv("/Users/christinacomo/Desktop/Spatial/Raw Data/Control1LGE.csv")
Midline <- read.csv("/Users/christinacomo/Desktop/Spatial/Raw Data/Control1Midline.csv")
Dorsal_Cortex <- read.csv("/Users/christinacomo/Desktop/Spatial/Raw Data/Control1DorsalCortex.csv")
Ventral_Cortex <- read.csv("/Users/christinacomo/Desktop/Spatial/Raw Data/Control1VentralCortex.csv")
```

```{r}
# rename columns in cortex and meninges dataframes for merging
names(MGE) <- c("barcodes", "region")
names(LGE) <- c("barcodes", "region")
names(Midline) <- c("barcodes", "region")
names(Dorsal_Cortex) <- c("barcodes", "region")
names(Ventral_Cortex) <- c("barcodes", "region")
```

```{r}
# merge cortex and meninges dataframes together
regions_of_interest <- rbind(MGE, LGE, Midline, Dorsal_Cortex, Ventral_Cortex) 
```


```{r}
# grab the spot barcodes to use for subsetting
spots.use <- regions_of_interest$barcodes
regions_of_interest <- regions_of_interest %>% column_to_rownames(var = "barcodes")
```

```{r}
# check dimensions of regions_of_interest and make sure it is the sum of cc_cortex and cc_meninges
dim(regions_of_interest)
```

```{r}
# now, subset the seurat object
SampleA <- subset(SampleA, cells = spots.use)
```

```{r}
subset_A <- AddMetaData(SampleA, metadata = regions_of_interest)
p1 <- SpatialDimPlot(subset_A)
p1
```


```{r}
SpatialFeaturePlot(subset_A, 
                   features = c("nCount_Spatial", 
                                "nFeature_Spatial", 
                                "percent_mito", 
                                "percent_hb"))
```

```{r}
saveRDS(subset_A, "/Users/christinacomo/OneDrive - The University of Colorado Denver/Spatial/RDS_files/SampleA_subsetwithbarcodes.RDS")
```

```{r}
subset_A <- readRDS("/Users/christinacomo/OneDrive - The University of Colorado Denver/Spatial/RDS_files/SampleA_subsetwithbarcodes.RDS")
subset_B <- readRDS("/Users/christinacomo/OneDrive - The University of Colorado Denver/Spatial/RDS_files/SampleB_subsetwithbarcodes.RDS")
subset_C <- readRDS("/Users/christinacomo/OneDrive - The University of Colorado Denver/Spatial/RDS_files/SampleC_subsetwithbarcodes.RDS")
subset_D <- readRDS("/Users/christinacomo/OneDrive - The University of Colorado Denver/Spatial/RDS_files/SampleD_subsetwithbarcodes.RDS")
```

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "/Users/christinacomo/OneDrive - The University of Colorado Denver/Spatial")
```


```{r}
setwd("/Users/christinacomo/OneDrive - The University of Colorado Denver/Spatial")
getwd()
```

## 2.1 Prepare data

```{r}
sample_id <- c("control1", "control2", "mutant1", "mutant2")
samples <- c()
spotfiles <- c()
imgs <- c()
json <- c()
for (sample in sample_id) {
  samples <- c(samples, paste("raw_data/", sample, "/filtered_feature_bc_matrix.h5", sep = ""))
  spotfiles <- c(spotfiles, paste("raw_data/", sample, "/spatial/tissue_positions_list.csv", sep = ""))
  imgs <- c(imgs, paste("raw_data/", sample, "/spatial/tissue_hires_image.png", sep = ""))
  json <- c(json, paste("raw_data/", sample, "/spatial/scalefactors_json.json", sep = ""))
}
infoTable <- data.frame(samples, spotfiles, imgs, json, sample_id)
```


```{r}
so <- InputFromTable(infotable = infoTable,
                      minUMICountsPerGene = 5,
                      minUMICountsPerSpot = 300,
                      platform =  "Visium")
```

```{r}
# convert sample_id to factor and reorder levels for plotting
so@meta.data$sample_id <- as.factor(so@meta.data$sample_id)
so@meta.data$sample_id <- factor(so@meta.data$sample_id, levels = c("control1", "control2", "mutant1", "mutant2"))
```

# 3 Quality Control

## 3.1 Mitochondrial content

We can explore mitochondrial transcript content in our data by identifying all mitochondrial genes (i.e., genes that have a "mt-" prefix) and then calculating the percentage of mitochondrial content per spot. This information is added to the *meta.data* of our Seurat object.

```{r}
mt.genes <- grep(pattern = "^mt-", x = rownames(so), value = TRUE)
so$percent.mt <- (Matrix::colSums(so@assays$RNA@counts[mt.genes, ])/Matrix::colSums(so@assays$RNA@counts))*100
```

Plot the percentage of mitochondrial content per spot per sample:

```{r, fig.height = 3, fig.width = 3}
ST.FeaturePlot(so, 
               features = "percent.mt", 
               palette = "Spectral",
               pt.size = 1.0, 
               ncol = 2, 
               label.by = "sample_id") +
  theme(plot.title = element_text(hjust = 0.5, 
                                  size = 18),
        text = element_text(size = 16))
```

Same information as above, but represented as a violin plot:

```{r}
VlnPlot(so, 
        features = "percent.mt",
        group.by = "sample_id",
        cols = custom_colors$discrete,
        pt.size = 0.1) +
  coord_flip() +
  NoLegend() +
  theme(plot.title = element_text(hjust = 0.5, 
                                  size = 18),
        text = element_text(size = 16))
```

## 3.2 nFeature (number of genes)

```{r, fig.height = 3, fig.width = 3}
ST.FeaturePlot(so, 
               features = "nFeature_RNA", 
               palette = "Spectral",
               pt.size = 1.0, 
               ncol = 2, 
               label.by = "sample_id") +
  theme(plot.title = element_text(hjust = 0.5, 
                                  size = 18),
        text = element_text(size = 16))
```

Same information as above, but represented as a violin plot:

```{r}
VlnPlot(so, 
        features = "nFeature_RNA",
        group.by = "sample_id",
        cols = custom_colors$discrete,
        pt.size = 0.1) +
  coord_flip() +
  NoLegend() +
  theme(plot.title = element_text(hjust = 0.5, 
                                  size = 18),
        text = element_text(size = 16))
```
## 3.3 nCount (total UMI count)

When we created our Seurat object we filtered out spots that had less than 300 total transcripts. Here, we'll plot the number of transcripts (UMI counts) per capture-spot for each sample.

```{r, fig.height = 3, fig.width = 3}
ST.FeaturePlot(so, 
               features = "nCount_RNA", 
               palette = "Spectral",
               pt.size = 1.0, 
               ncol = 2, 
               label.by = "sample_id") +
  theme(plot.title = element_text(hjust = 0.5, size = 18),
        text = element_text(size = 16))
```

Same information as above, but represented as a violin plot:

```{r}
VlnPlot(so, 
        features = "nCount_RNA",
        group.by = "sample_id",
        cols = custom_colors$discrete,
        pt.size = 0.1) +
  coord_flip() +
  NoLegend() +
  theme(plot.title = element_text(hjust = 0.5, 
                                  size = 18),
        text = element_text(size = 16))
```
# 4 Image Processing

## 4.1 Loading and handling images

Now that the Seurat object is created and filtered, we can load the H&E images from the file paths provided in the *infoTable*. 

```{r}
so <- LoadImages(so, time.resolve = TRUE, verbose = TRUE)
```

Take a quick look at the raw (unprocessed) sections:

```{r}
ImagePlot(so, method = "raster", type = "raw")
```

## 4.2 Mask Images

Let's mask the background of the H&E images so that we can rotate them (i.e., it is easier to look at the sections when the dorsal-ventral axis is oriented top-to-bottom versus right-to-left).

```{r}
so <- MaskImages(object = so)
```

Plot the masked images:

```{r}
ImagePlot(so, method = "raster", type = "masked")
```
## 4.3 Transform images

Here we'll apply rigid transformations to our images so that they are in a dorsal-ventral orientation from top-to-bottom (versus right-to-left).

```{r}
transforms <- list("1" = list("angle" = 270),
                   "2" = list("angle" = 270),
                   "3" = list("angle" = 270),
                   "4" = list("angle" = 270))
so <- WarpImages(so, transforms)
```

Plot rotated images:

```{r}
ImagePlot(so, method = "raster", type = "processed")
```
```{r, fig.height = 3, fig.width = 3}
FeatureOverlay(so, 
               features = "Vim", 
               sampleids = 1:4,
               cols = c("lightgray", "mistyrose", "red", "dark red", "black"),
               pt.size = 1.0, 
               ncol = 2, 
               label.by = "sample_id",
               type = "raw")
```


```{r, fig.height = 3, fig.width = 3}
FeatureOverlay(so, 
               features = "Vim", 
               sampleids = 1:4,
               cols = c("lightgray", "mistyrose", "red", "dark red", "black"),
               pt.size = 1.0, 
               ncol = 2, 
               label.by = "sample_id",
               type = "processed")
```

# 5 Normalization

The [spatial vignette](https://satijalab.org/seurat/articles/spatial_vignette.html) from Seurat recommends using sctransform to normalize ST data, so we will use sctransform to normalize our data. Note that `SCTransform()` replaces `NormalizeData()`, `ScaleData()`, and `FindVariableFeatures()`.

```{r}
so <- SCTransform(so)
```
## 6.1 Clustering

We will use the SCT assay of our Seurat object for clustering.

```{r}
so <- RunPCA(so, assay = "SCT", verbose = FALSE)
so <- FindNeighbors(so, reduction = "pca", dims = 1:50, verbose = FALSE)
so <- FindClusters(so, verbose = FALSE)
```

### 6.1.1 Plot the clusters spatially

Plot all clusters together spatially for each sample.

```{r, fig.height = 3, fig.width = 3}
ST.FeaturePlot(object = so, 
               features = "seurat_clusters", 
               pt.size = 1, 
               ncol = 2,
               show.sb = FALSE,
               label.by = "sample_id")
```

#### 6.1.2.1 control1

```{r}
# grab cluster color assignments from previous plot for consistency
cluster_colors <- c("#FFC312", "#C4E538", "#12CBC4", "#FDA7DF", "#ED4C67", "#F79F1F", "#A3CB38", "#1289A7", "#D980FA", "#B53471", "#EE5A24", "#009432", "#0652DD", "#9980FA", "#833471", "#EA2027", "#006266", "#1B1464", "#c9cece")
names(cluster_colors) <- c("0", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15", "16", "17", "-")
```

```{r, fig.height = 3, fig.width = 3}
ST.FeaturePlot(object = so, 
               features = "seurat_clusters",
               pt.size = 1, 
               split.labels = T, 
               indices = 1, 
               show.sb = FALSE, 
               ncol = 5,
               cols = cluster_colors)
```

#### 6.1.2.2 control2

```{r, fig.height = 3, fig.width = 3}
ST.FeaturePlot(object = so, 
               features = "seurat_clusters",
               pt.size = 1, 
               split.labels = T, 
               indices = 2, 
               show.sb = FALSE, 
               ncol = 5,
               cols = cluster_colors)
```

#### 6.1.2.3 mutant1

```{r, fig.height = 3, fig.width = 3}
ST.FeaturePlot(object = so, 
               features = "seurat_clusters",
               pt.size = 1, 
               split.labels = T, 
               indices = 3, 
               show.sb = FALSE, 
               ncol = 5,
               cols = cluster_colors)
```

#### 6.1.2.4 mutant2

```{r, fig.height = 3, fig.width = 3}
ST.FeaturePlot(object = so, 
               features = "seurat_clusters",
               pt.size = 1, 
               split.labels = T, 
               indices = 4, 
               show.sb = FALSE, 
               ncol = 5,
               cols = cluster_colors)
```

## 6.2 Dimensional reduction

### 6.2.1 Calculate the UMAP

```{r}
so <- RunUMAP(so, reduction = "pca", dims = 1:50)
```

### 6.2.2 Plot the UMAP

#### 6.2.2.1 By sample and by cluster

```{r}
plot_umap_by_sample <- bind_cols(so@meta.data, as.data.frame(so@reductions$umap@cell.embeddings)) %>%
  ggplot(aes(UMAP_1, UMAP_2, color = sample_id)) +
  geom_point(size = 0.2) +
  theme_bw() +
  scale_color_manual(values = custom_colors$discrete) +
  labs(color = 'Sample') +
  guides(colour = guide_legend(override.aes = list(size = 2))) +
  theme(legend.position = 'left') +
  coord_fixed() +
  annotate(
    geom = 'text', x = Inf, y = -Inf,
    label = paste0('n = ', format(nrow(so@meta.data), big.mark = ',', trim = TRUE)),
    vjust = -1.5, hjust = 1.25, color = 'black', size = 2.5
  )
plot_umap_by_cluster <- bind_cols(so@meta.data, as.data.frame(so@reductions$umap@cell.embeddings)) %>%
  ggplot(aes(UMAP_1, UMAP_2, color = seurat_clusters)) +
  geom_point(size = 0.2) +
  theme_bw() +
  scale_color_manual(values = custom_colors$discrete) +
  labs(color = 'Cluster') +
  guides(colour = guide_legend(override.aes = list(size = 2))) +
  theme(legend.position = 'right') +
  coord_fixed() +
  annotate(
    geom = 'text', x = Inf, y = -Inf,
    label = paste0('n = ', format(nrow(so@meta.data), big.mark = ',', trim = TRUE)),
    vjust = -1.5, hjust = 1.25, color = 'black', size = 2.5
  )

plot_umap_by_sample 
 
```

# 8 Save Data

```{r}
saveRDS(so, "/Users/christinacomo/OneDrive - The University of Colorado Denver/Spatial/RDS_files/so_all.rds")
```







